---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# likertr

<!-- badges: start -->
<!-- badges: end -->

Likertr provides a comprehensive workflow for a full-scale analysis of Likert-scale data. It provides:

1. Data Preparation and Cleaning
  (a) Dealing with "neutral" or "do not know" or "do not wish to answer".
  (b) Center weighting or setting to NA.

2. Reliability Measures and Structure Information
  (a) Cronbach’s alpha and McDonald’s omega.
  (b) Exploratory Factor Analysis.
    i. Kaiser-Meyer-Olkin (KMO): A test to see if the items share enough variance
    ii. Bartlett’s Test of Sphericity: A test to confirm the correlation matrix isn’t just random noise
    iii. Polychoric Correlation Matrix: This treats the Likert categories as ordinal rather than continuous.
  (c) Relative Importance Index (RII) - a way to rank the importance of survey questions against each other.

3. Visualization
  (a) Diverging stacked bar charts.
  (b) Heat maps for correlations.
  (c) Comparison plots (group to group).
  
4. Inference and Reporting
  (a) Nonparametric testing with Mann-Whitney U or Kruskal-Wallis.
  (b) Effect size calculation (Cliff’s Delta or r).

## Installation

You can install the development version of likertr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("laurenlaturner/likertr")
```

## Example Workflow

Three main functions in Likertr provide all required functionality. The function "likertr()" is used to create a likertr object that can then be passed to summary() and plot() generics.

```{r example}
library(likertr)

load("data/data.rda")

analysis <- likertr(data)
summary(analysis)
plot(analysis)

```


