---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# likertr

<!-- badges: start -->
<!-- badges: end -->

`likertr` provides a comprehensive workflow for a full-scale analysis of Likert-scale data. The package facilitates the following processes:

## 1. Data Preparation and Cleaning
* **Handling specific responses:** Dealing with "neutral," "do not know," or "do not wish to answer."
* **Recoding:** Center weighting or setting values to `NA`.

## 2. Reliability Measures and Structure Information
* **Consistency Metrics:** Cronbach’s alpha and McDonald’s omega.
* **Exploratory Factor Analysis (EFA):**
    * **Kaiser-Meyer-Olkin (KMO):** A test to see if the items share enough variance.
    * **Bartlett’s Test of Sphericity:** A test to confirm the correlation matrix isn’t just random noise.
    * **Polychoric Correlation Matrix:** Treats Likert categories as ordinal rather than continuous.
* **Relative Importance Index (RII):** A method to rank the importance of survey questions against each other.

## 3. Visualization
* **Diverging stacked bar charts:** The standard for Likert scale visualization.
* **Heat maps:** Specifically for correlation matrices.
* **Comparison plots:** Useful for group-to-group analysis.

## 4. Inference and Reporting
* **Nonparametric testing:** Utilizing Mann-Whitney U or Kruskal-Wallis tests.
* **Effect size calculation:** Providing Cliff’s Delta or $r$ for robust interpretation.
## Installation

You can install the development version of likertr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("laurenlaturner/likertr")
```
## Example Workflow

Three main functions in Likertr provide all required functionality. The function "likertr()" is used to create a likertr object that can then be passed to summary() and plot() generics.

```{r example}
library(likertr)

load("data/data.rda")

analysis <- likertr(data)
summary(analysis)
plot(analysis)

```


