---
title: "Getting Started with likertr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with likertr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(likertr)

```
likertr is a lightweight package that can analyze your likert scale data in one simple step! Here is a quick tutorial on what inputs you should provide and what output will come back out.

1. Summary Command
likertr comes with example data on student stress levels. Lets look at the data:
```{r, include = TRUE}
library(likertr)
head(data)

```

Lots to summarize! The only required argument of the summary function is just the dataset. 

```{r, include = TRUE}
#summary(data)
```

Likertr also has custom plots that come with summary. 

```{r, include = TRUE}
#layout(matrix(1:6,nrow = 3, ncol =2, byrow = TRUE))
#plot(likertr)
```


If you want more control over the analysis that is happening in likertr, other arguments can be specified.

Likertr is designed to handle 'NA' values effectively. Default handling will drop observations when NA values interfere with analysis. The argument na_drop option "neutral" can also be specified, replacing NA values with the midpoint.

ipsatize_decision allows you to center all responses around the respondent's mean response.

small_n_drop lets you drop questions with less than 20 respondent's.

groups allows you to specify question groups that you want to compare to each other, both for calculating Cronbach's alpha and calculating Cliff's Delta Value for correlation between responses.
```{r, include = TRUE}
# Compare whether students depression responses are correlated to anxiety responses.

```

If you have a categorical variable you want to split your data on (or perhaps a likert scale question that has large number of responses), inference can be done on whether those groups are significantly different from each other.






