---
title: "Getting Started with likertr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with likertr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  # = "80%",
  fig.align = "center"
)
```
Welcome friend. If you want to see our github, check out the link below!
github link: https://github.com/laurenlaturner/likertr.git
```{r setup}

library(likertr)
```
likertr is a lightweight package that can analyze your likert scale data in one simple step! Here is a quick tutorial on what inputs you should provide and what output will come back out.


likertr comes with example data on student stress and depression levels. We'll use 5 questions about depressiona and anxiety as variables for the demonstration. Lets preview the data:
```{r, include = TRUE}
library(likertr)
small_data <- data[c(1, 3:7, 11:15)]
str(small_data)

```

Lots of analysis to do! The most basic way to use this package is to put a data set into the likertr function and let the defaults do their work.

```{r, include = TRUE}
likert_analysis <- likertr(small_data)
str(likert_analysis)
```

The likertr function returns a custom s3 object with summary and plotting functions

```{r, inclue = TRUE}
summary(likert_analysis)
```




```{r, include = TRUE}
plot(likert_analysis)
```


If you want more control over the analysis that is happening in likertr, other arguments can be specified, including handling of NA values, scaling, and checking for sufficient number of observations.

```{r, echo = TRUE, results = 'hide'}
# Perform likert analysis replacing NA values with the midpoint, 
# center respondents answers around their mean answer, and drop columns with 
# less than 20 sample sizes.
clean_analysis <- likertr(small_data, 
                          na_drop = "neutral", 
                          ipsatize_decision = TRUE,
                          small_n_drop = TRUE)
```



likertr does exploratory factor analysis. The number of underlying factors can be specified, otherwise parallel analysis will decide the ideal number of factors.

The groups argument can be used to group similar likert questions together to find Cronbach's Alpha, a measure of consistency of the data. The groups vector should be the same length as the number of questions, with each integer specifying which group the question falls in. If specified the groups will also be compared against each other in the inference step to see if the responses between groups are significantly different from each other.

A categorical variable can also be specified in order to perform inference on whether answers are significantly different across groups.


```{r, include = TRUE}
# Perform the likert analysis with two underlying factors,
# the anxiety and depression questions split into groups for similarity,
# and the gender categorical variable used for inference.
likert_analysis <- likertr(small_data,
                           n_fact = 2,
                           groups = c(rep(1, 5), rep(2, 5)),
                           category = 1)
summary(likert_analysis)
```





